---
params:
  authors: 'Ben Anderson'
  title: 'On the timing of earthquakes (and their magnitude)'
  subtitle: 'Using Aotearoa New Zealand GeoNet data'
title: '`r paste0(params$title)`'
subtitle: '`r paste0(params$subtitle)`'
author: '`r paste0(params$authors)` (Contact: b.anderson@soton.ac.uk, `@dataknut`)'
date: 'Last run at: `r Sys.time()`'
always_allow_html: yes
output:
  bookdown::html_document2:
    code_folding: hide
    fig_caption: yes
    number_sections: yes
    self_contained: no
    toc: yes
    toc_depth: 2
    toc_float: yes
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(skimr)
library(ggplot2)

# get the data back from wherever drake put it
dt <- drake::readd(drakeData)
  
```

# Background

We had a bit or rock n roll on [Monday night](https://www.geonet.org.nz/earthquake/2019p603895). So the question naturally arose: `are earthquakes more likely to occur at night?`.

You might [think not](https://www.quora.com/Why-do-earthquakes-happen-at-night-most-of-the-time), I couldn't possibly comment.

Luckily in Aotearoa New Zealand [GeoNet](https://www.geonet.org.nz/) makes answering the question very easy. Just [download](https://quakesearch.geonet.org.nz/) the most recent year's worth of earthquake records and off you go...

# Data

We downloaded all earthquake records from `r startDate` to `r endDate`. This is a grand total of `r nrow(dt)` records.

Not all the events are actually earthquakes. As Table \@ref(tab:eventsTab) shows, there are explosions, quarry blasts etc.

```{r eventsTab}
t <- cube(dt, j = list("min magnitude" = min(magnitude),
                       "mean magnitude" = mean(magnitude),
                       "max magnitude" = max(magnitude), 
                       "n Obs" = .N), by = c("eventtype"))

kableExtra::kable(t[!is.na(eventtype)], digits = 2,
                  caption = "Mean magnitude and number of events") %>%
  kable_styling()

dt <- dt[eventtype %like% "earthquake"]
dt <- dt[, hms := hms::as.hms(origintime)]
dt <- dt[, halfHour := hms::trunc_hms(hms, 60*30)]
```

As we can see there are some (`r nrow(dt[magnitude <= 0])`) records with magnitude < 0?

# Q1: Are earthquakes more likely to occur at night?

To answer this we plot all earthquakes where magnitude > 0.

```{r allPoint, fig.cap="Number of earthquakes by time of day"}
plotDT <- dt[magnitude > 0, .(nObs = .N,
                 meanMag = mean(magnitude)), keyby = .(halfHour)]

myCaption <- paste0("Source: NZ GeoNet (https://quakesearch.geonet.org.nz/)\nAll earthquakes ",
                    min(lubridate::date(dt$origintime)), " to ", max(lubridate::date(dt$origintime))
                    )

p <- ggplot2::ggplot(plotDT, aes(x = halfHour, 
                                 y = nObs, 
                                 colour = round(meanMag,1)
                                 )
                     ) +
  geom_point() +
  labs(x = "Time of Day (half hours)",
       y = "Number of Observations",
       caption = paste0(myCaption,
                        "\nAll magnitude > 0 (", nrow(dt[magnitude > 0]), ")")) +
  scale_color_continuous(low = "green", high = "red", name = "Mean magnitude") +
  theme(legend.position="bottom")

p
```

So. Figure \@ref(fig:allPoint) suggests more are detected at night. Really? But there is a little clue in the colour of the points - the mean magnitude is higher during the day. This also seems unlikely.

# Q2: Are large earthquakes more likely to occur during the day?

To answer this we plot all earthquakes where magnitude > 3.

```{r allPoint3, fig.cap="Number of earthquakes by time of day (magnitude > 3)"}
plotDT <- dt[magnitude > 3, .(nObs = .N,
                 meanMag = mean(magnitude)), keyby = .(halfHour)]

p <- ggplot2::ggplot(plotDT, aes(x = halfHour, 
                                 y = nObs, 
                                 colour = round(meanMag,1)
                                 )
                     ) +
  geom_point() +
  labs(x = "Time of Day (half hours)",
       y = "Number of Observations",
       caption = paste0(myCaption,
                        "\nAll magnitude > 3 (", nrow(dt[magnitude > 3]), ")")
                        ) +
  scale_color_continuous(low = "green", high = "red", name = "Mean magnitude") +
  theme(legend.position="bottom")

p
```

This suggests that it is just easier to detect lower magnitude earthquakes at night when people are asleep, there are no trucks on the roads etc.

# Q3: Are stronger earthquakes more likely during the day?

Just to check:

```{r magBox, fig.cap="Mean magnitude of earthquakes by time of day"}
p <- ggplot2::ggplot(dt[magnitude > 0], aes(x = halfHour, y = magnitude, group = halfHour)) +
  geom_boxplot() +
  labs(x = "Time of Day",
       y = "Magnitude",
       caption = paste0(myCaption,
                        "\nAll magnitude > 0 (", nrow(dt[magnitude > 0]), ")")
                        ) +
  theme(legend.position="bottom")
p
```

No they are not.

# Data Annex

Original data as loaded from `r dataURL`:

```{r dataSkim}
skimr::skim(df) # original data
```

# R packages used

 * base R [@baseR]
 * data.table [@data.table] - for data munching
 * drake [@drake] - for data pre-loading and caching
 * ggplot2 [@ggplot2] - for plots
 * hms [@hms] - for time munching
 * kableExtra [@kableExtra] - for nice tables
 * lubridate [@lubridate] - for dateTime munching
 * readr [@readr] - for data download & parsing
 * skimr [@skimr] - for data description
 
# References